MolProbity's Ultimate Rotamer-Library Distributions for Model Validation

### 2016-02-19 response_2_reviwers.txt

"We therefore call these datasets the Top8000, as successors to the Top500.”

* Section 5.1 should be a single well organized paragraph, not a collection of 5 single sentence paragraphs.  I’m also not completely sure this section (and the designation of ultimate) is completely necessary. It is quite clear you have made a contribution, it is perhaps less clear that we can be absolutely sure we will not want to iterate rotamer library development one or more times after this.

-->  It would be extremely difficult to make this section readable and its logic understandable as a single paragraph.  Instead, we have made it an explicit list of separate arguments and have rewritten it for better clarity and continuity, which is indeed quite helpful:

"Calling this set of rotamer-library distributions ”ultimate” is a claim that requires both explanation and justification. 16 years after our ”penultimate” rotamers, we had to confront the issue of whether or not MolProbity’s line of validation-focused rotamer distributions had reached its ultimate stage. We decided that indeed it had, from three separate lines of argument, reflected in the wording of this work’s title.

(1) We narrowly claim it as ultimate for the multi-dimensional χ distributions used to validate sidechain models, such as done by the MolProbity website, in PHENIX, or by the wwPDB. In other words, this dataset is sufficient and will remain sufficient for sidechain validation in the current mode. The rotamers presented here are of course not necessarily ultimate or even appropriate for other purposes –including new validation methods that do not assume as much independence between the backbone and sidechain as current methods do.

(2) The changes from penultimate to ultimate distributions are fairly minor in quantitative terms, even after 16 years. The importance of the new system lies in its ability to support a 3-level evaluation, to identify very rare rotamers, and to specify rotamer-dependent bond-angle deviations. A million reliably-modeled residues of data now provide fully adequate sampling for the basic task of robustly locating smooth contours that separate favored (98%), allowed, and outlier regions. The fundamental χ distributions would be unlikely to change much at all if they were redetermined in the future, and we plan to keep them freely available on GitHub for testing, for use within other software systems, and for spinoffs such as approximation by differentiable functions.

(3) What we do expect will improve in the future is a redefinition of conformational validation, enabled by expansion of data and compute power, and driven by the needs of structure determination at lower resolution. Rather than doing separate Ramachandran and rotamer evaluation, we should move toward analyzing all backbone and sidechain torsional dimensions together [8], including allowance for the influence of secondary structure and local motifs. Thus our ”ultimate” claim asserts the position that MolProbity’s type of rotamer evaluation should no longer be updated but should evolve into something that treats covariation simultaneously."

### 2016-02-20 abstract.tex
Group revisions

Here we describe the updated MolProbity rotamer-library distributions derived from an order-of-magnitude larger and more stringently quality-filtered dataset of about 8000 (vs. 500) protein chains, and we explain the resulting changes and improvements to model validation as seen by users. To include only sidechains with satisfactory justification for their given conformation, we added residue-specific filters for electron-density value and model-to-density fit.  The combined new protocol retains a million residues of data, while cleaning up \textcolor{changecolor}{false-positive} noise in the multi-$\chi$ \textcolor{changecolor}{datapoint} distributions.  It enables clean characterization of conformational clusters nearly 1000-fold less frequent than the most common ones.  We describe examples of local interactions that favor these rare conformations, including the role of authentic covalent bond-angle deviations in enabling presumably strained sidechain conformations. Further, along with \textit{favored} and \textit{outlier}, an \textit{allowed} category (0.3\% to 2.0\% occurrence in reference data) has been added, analogous to Ramachandran validation categories. The new rotamer distributions are used for current rotamer validation in MolProbity and \textit{PHENIX}, and for rotamer choice in \textit{PHENIX} model-building and refinement. The multi-dimensional $\chi$ distributions and Top8000 reference dataset are available on GitHub. These rotamers are termed "ultimate" because data sampling and quality are now fully adequate for this task, and also because we believe the future of conformational validation should integrate sidechain and backbone \textcolor{changecolor}{criteria}.

>>>>

Here we describe the updated MolProbity rotamer-library distributions derived from an order-of-magnitude larger and more stringently quality-filtered dataset of about 8000 (vs. 500) protein chains, and we explain the resulting changes and improvements to model validation as seen by users. To include only sidechains with satisfactory justification for their given conformation, we added residue-specific filters for electron-density value and model-to-density fit.  The combined new protocol retains a million residues of data, while cleaning up \textcolor{changecolor}{false-positive} noise in the multi-$\chi$ \textcolor{changecolor}{datapoint} distributions.  It enables \textcolor{changecolor}{unambiguous} characterization of conformational clusters nearly 1000-fold less frequent than the most common ones.  We describe examples of local interactions that favor these rare conformations, including the role of authentic covalent bond-angle deviations in enabling presumably strained sidechain conformations. Further, along with \textit{favored} and \textit{outlier}, an \textit{allowed} category (0.3\% to 2.0\% occurrence in reference data) has been added, analogous to Ramachandran validation categories. The new rotamer distributions are used for current rotamer validation in MolProbity and \textit{PHENIX}, and for rotamer choice in \textit{PHENIX} model-building and refinement. The multi-dimensional $\chi$ distributions and Top8000 reference dataset are freely available on GitHub. These rotamers are termed "ultimate" because data sampling and quality are now fully adequate for this task, and also because we believe the future of conformational validation should integrate sidechain and backbone \textcolor{changecolor}{criteria, as opposed to the current independent treatment}.

### 2016-02-20 discussion.tex
Group revisions

\textcolor{changecolor}{(1) 
We narrowly claim it as ultimate for the multi-dimensional $\chi$ distributions used to validate sidechain models, such as done by the MolProbity website, in \textit{PHENIX}, or by the wwPDB. In other words, this dataset is sufficient and will remain sufficient for sidechain validation in the current mode. The rotamers presented here are of course not necessarily ultimate or even appropriate for other purposes --including new validation methods that do not assume as much independence between the backbone and sidechain as current methods do.}

@@

\textcolor{changecolor}{(3)  W}hat we do expect will improve in the future is a redefinition of conformational validation, enabled by expansion of data and compute power, and \textcolor{changecolor}{driven by} the needs of structure determination at lower resolution. Rather than doing separate Ramachandran and rotamer evaluation, we should move toward analyzing all backbone and sidechain torsional dimensions together \cite{JSR_theplot_2013}, including allowance for the influence of secondary structure and local motifs. Thus our "ultimate" claim asserts the position that \textcolor{changecolor}{MolProbity's type of }rotamer evaluation should no longer be updated but should evolve into something \textcolor{changecolor}{that treats covariation simultaneously}.

@@

Sometimes we get disturbing reports that people are trying to achieve a sort of "MolProbity Nirvana" by attempting to eliminate every single outlier. To approach eliminating all clashes is a worthy goal: in principle the clash target is zero, as sterics do not allow serious atomic overlaps. However, at high resolution a zero score would require the difficult reconciliation of occupancies for alternate conformations including waters, and in any case a few puzzles often remain. Another important point to keep in mind is that the primary purpose and usefulness of model validation is to help diagnose and correct places where the conformation has been fit in the wrong local energy well. Such corrections often matter to biological interpretation, and will be stable to further refinement. In contrast, a small shift across the border into the allowed region will often shift right back again, and is not a very meaningful improvement.

However, the target is very definitely \textit{not} zero for \textcolor{changecolor}{Ramachandran and rotamer validation. The idea that \textit{every single} torsion} needs to be allowed is a misunderstanding. As outlined here, validation of torsion angles relies on contoured conformational distributions from quality-filtered datasets. The outer contour defines an outlier cutoff, meaning that even the filtered reference data has outliers but \textit{not} that those conformations are wrong. There is room for valid torsion outliers, more so for sidechain rotamers than for backbone Ramachandran values. Even in 2005 the curve of rotamer-outlier percentage, fit as a function of resolution, was found to asymptote at 0.5\% \cite{Arendall2005}. For quality-filtered data the rotamer outlier cutoff is now 0.3\%, meaning that 3 out of 1,000 residues are expected to be valid rotamer outliers. For Ramachandran the outlier cutoff is 0.05\%, so only a very low 5 out of 10,000 residues -- but not zero -- are expected to be valid Ramachandran outliers. In either case, to be accepted as valid, an outlier should have clear electron density to support its occurrence, and should have either H-bonds or tight packing to hold it in the presumably quite strained conformation. This is exemplified by the Asn rotamer outliers in Figure \ref{fig:AsnOutliers}, which each have an eclipsed $\chi$1 but are validated by excellent density and three sidechain H-bonds.  Valid outliers are more likely to have been selected and maintained by evolution if there is a functional need of some sort (folding, catalysis, binding) for that specific conformation; therefore such cases will probably be of great interest and will reward as well as require detailed examination.

>>>>

\textcolor{changecolor}{(1) 
We claim it as ultimate only for the multi-dimensional $\chi$ distributions used to validate sidechain models, such as done by the MolProbity website and in \textit{PHENIX}. The rotamers presented here are of course not necessarily ultimate or even appropriate for other purposes.}

@@

\textcolor{changecolor}{(3)  W}hat we do expect will improve in the future is a redefinition of conformational validation, enabled by expansion of data and computing power, and \textcolor{changecolor}{driven by} the needs of structure determination at lower resolution. Rather than doing separate Ramachandran and rotamer evaluation, we should move toward analyzing all backbone and sidechain torsional dimensions together \cite{JSR_theplot_2013}, including allowance for the influence of secondary structure and local motifs. Thus our "ultimate" claim asserts the position that \textcolor{changecolor}{MolProbity's type of }rotamer evaluation should no longer be updated but should evolve into something \textcolor{changecolor}{that treats covariation simultaneously}.

@@

Sometimes we get disturbing reports that people are trying to achieve a sort of "MolProbity Nirvana" by attempting to eliminate every single outlier. To approach eliminating all clashes is a worthy goal: in principle the clash target is zero, as sterics do not allow serious atomic overlaps. In practice however, (a) all-atom contact analysis includes approximations (e.g. spherical atoms); (b) at high resolution a zero score would require the difficult reconciliation of occupancies for alternate conformations including waters; and (c) usually a few puzzles remain. An important point to keep in mind is that the primary purpose and usefulness of model validation is to help diagnose and correct places where the conformation has been fit in the wrong local energy well. Such corrections often matter to biological interpretation, and will be stable to further refinement. In contrast, a small shift across the border into the allowed region will often shift right back again, and is not a very meaningful improvement.

However, the target is very definitely \textit{not} zero for \textcolor{changecolor}{Ramachandran and rotamer validation. The idea that \textit{every single} torsion} needs to be allowed is a misunderstanding. As outlined here, validation of torsion angles relies on contoured conformational distributions from quality-filtered datasets. The outer contour defines an outlier cutoff, meaning that even the filtered reference data has outliers but \textit{not} that those conformations are wrong. There is room for valid torsion outliers, more so for sidechain rotamers than for backbone Ramachandran values. Even in 2005 the curve of rotamer-outlier percentage, fit as a function of resolution, was found to asymptote at 0.5\% \cite{Arendall2005}. For quality-filtered data the rotamer outlier cutoff is now 0.3\%, meaning that 3 out of 1,000 residues are expected to be valid rotamer outliers. For Ramachandran the outlier cutoff is 0.05\%, so only a very low 5 out of 10,000 residues -- but not zero -- are expected to be valid Ramachandran outliers. In either case, to be accepted as valid, an outlier should have clear electron density to support its occurrence, and should have either H-bonds or tight packing to hold it in the presumably quite strained conformation. This is exemplified by the Asn rotamer outliers in Figure \ref{fig:AsnOutliers}, which each have an eclipsed $\chi$1 but are validated by excellent density and three sidechain H-bonds.  Valid outliers are more likely to have been selected and maintained by evolution if there is a functional need of some sort (folding, catalysis, binding) for that specific conformation; therefore such cases will probably be of interest and will reward as well as require detailed examination.

### 2016-02-20 introduction.tex
Group revisions

\section{Introduction}
Protein sidechains take on preferred conformations which fall into distinct local energy minima known as rotamers, defined by the set of sidechain dihedral ($\chi$) angles. For tetrahedral geometry, $\chi$ values fall into three discrete ranges: \textbf{p} (plus, centered near +60$\degree$), \textbf{t} (trans, centered near 180$\degree$), and \textbf{m} (minus, centered near -60$\degree$)\textcolor{changecolor}{, as named in\cite{lovell2000penultimate}\footnote{\textcolor{changecolor}{The p, t, m nomenclature was adopted in \cite{lovell2000penultimate} and in MolProbity \cite{Davis2004}, to give a single letter for use in rotamer strings, and because the more common g+, t, g- terminology was, in 2000, still being assigned to opposite conformations (see discussion in \cite{lovell2000penultimate}).}}}. These correspond to low-energy staggered conformations expected between sp$^{3}$ hybridized atoms \cite{Eyring1932}. sp$^{3}$ to sp$^{2}$ bonds (tetrahedral to planar geometries) have more complex and much broader distributions. Overall rotamer conformations, however, are not simply the product of their individual $\chi$ distributions, since wider steric and other atomic interactions influence the preferred, or even the possible, combinations.

@@

Different rotamer libraries represent the allowed variability around each central conformation in one of three ways. Early rotamer libraries simply provided the mean value and some estimate of allowable range for all $\chi$ angles (or, often, just for $\chi$1 and $\chi$2) of each identified rotamer for each side-chain type \cite{Ponder1987, Tuffery1991,Schrauber1993}.  A user would either simply use the mean-value conformation, or else would optimize it manually or computationally within the allowable range. Because modeling the lowest-energy conformation fails to capture allowed variation and further minimization is computationally expensive, design methods expanded to employ "grid" libraries of arbitrarily-spaced, discrete sample points in $\chi$ space around the low-energy mean of each rotamer, which allowed development of the influential Dead-End Elimination method in protein design \cite{DeMaeyer1997,Gainza2013}.  A third type of rotamer software \cite{Dunbrack1997, Chen:2010kx} evaluates a given sidechain conformation by its position in a multi-dimensional probability distribution. Early such distributions were binned (often at $\ge$ 10 $\degree$) \cite{Laskowski:gl0276}, but recent ones use smooth contour surfaces, scored by what percent of the reference data lies outside that contour \cite{lovell2000penultimate, Read2011}. \textcolor{changecolor}{Design libraries and validation libraries focus on two distinct areas of the reference data distributions. While design and prediction are primarily concerned with statistics and cluster shapes inside the low-energy wells \cite{Dunbrack1997}, validation is primarily concerned with robustly identifying the outliers beyond the edges of those wells. Such outliers are usually wrong but sometimes valid and interesting, so are always worth examining \cite{Richardson2013}.}

@@

Previously our group developed the "penultimate" rotamer library, using the quality-filtered Top240 PDBs \cite{lovell2000penultimate}. Soon afterward we updated the library using our Top500 dataset \cite{Lovell:2003uq}. This library used many file-level filters such as requiring $\le$ 1.8 resolution, a clashscore $<$ 30 \cite{Word1999}, and few backbone bond-angle outliers from Engh and Huber standards \cite{Engh1991}, \textcolor{changecolor}{and when the file contained multiple identical chains, used only the best-ordered one}. The residue-level filters required a B factor $<$ 40 for all atoms in the residue and the absence of serious steric clashes, defined as hydrogen-aware atomic overlaps $\ge$ 0.4 \AA \cite{Word1999}. These filters were used in order to eliminate residues with questionable justification for their given conformation, thereby increasing reliability. Specifically, the B-factor filter (the best metric available before mandatory data deposition) was meant to eliminate residues with poor electron density or other local uncertainty. However, as we realized and as Shapovalov and Dunbrack later clearly demonstrated \cite{Shapovalov:2007}, in many structures the B factors are a poor indicator of good density, primarily because they are often restrained to change only modestly between adjacent, covalently bonded atoms. Since 2008, when the wwPDB started requiring structure-factor data with all depositions, it has become feasible to do routine analysis of electron density directly.

>>>>

\section{Introduction}
Protein sidechains take on preferred conformations which fall into distinct local energy minima known as rotamers, defined by the set of sidechain dihedral ($\chi$) angles. For tetrahedral geometry, $\chi$ values fall into three discrete ranges: \textbf{p} (plus, centered near +60$\degree$), \textbf{t} (trans, centered near 180$\degree$), and \textbf{m} (minus, centered near -60$\degree$)\textcolor{changecolor}{, as named in\cite{lovell2000penultimate}\footnote{\textcolor{changecolor}{The p, t, m nomenclature was adopted in \cite{lovell2000penultimate} and in MolProbity \cite{Davis2004}, to give a single letter for use in rotamer strings, and because the more common g+, t, g- terminology was, in 2000, still being assigned inconsistently (see discussion in \cite{lovell2000penultimate}).}}}. These correspond to low-energy staggered conformations expected between sp$^{3}$ hybridized atoms \cite{Eyring1932}. sp$^{3}$ to sp$^{2}$ bonds (tetrahedral to planar geometries) have more complex and much broader distributions. Overall rotamer conformations, however, are not simply the product of their individual $\chi$ distributions, since wider steric and other atomic interactions influence the preferred, or even the possible, combinations.

@@

Different rotamer libraries represent the allowed variability around each central conformation in one of three ways. Early rotamer libraries simply provided the mean value and some estimate of allowable range for all $\chi$ angles (or, often, just for $\chi$1 and $\chi$2) of each identified rotamer for each side-chain type \cite{Ponder1987, Tuffery1991,Schrauber1993}.  A user would either simply use the mean-value conformation, or else would optimize it manually or computationally within the allowable range. Because modeling the lowest-energy conformation fails to capture allowed variation and further minimization is computationally expensive, design methods expanded to employ "grid" libraries of arbitrarily-spaced, discrete sample points in $\chi$ space around the low-energy mean of each rotamer, which allowed development of the influential Dead-End Elimination method in protein design \cite{DeMaeyer1997,Gainza2013}.  A third type of rotamer software \cite{Dunbrack1997, Chen:2010kx} evaluates a given sidechain conformation by its position in a multi-dimensional probability distribution. Early such distributions were binned (often at $\ge$ 10 $\degree$) \cite{Laskowski:gl0276}, but recent ones use smooth contour surfaces, scored by what percent of the reference data lies outside that contour \cite{lovell2000penultimate, Read2011}. \textcolor{changecolor}{Design libraries and validation libraries focus on two distinct areas of the reference data distributions. While design and prediction are primarily concerned with statistics and cluster shapes inside the low-energy wells \cite{Dunbrack1997}, validation is primarily concerned with robustly identifying the outliers beyond the edges of those wells. Such outliers are usually wrong but sometimes valid and interesting, and so are always worth examining \cite{Richardson2013}.}

@@

Previously our group developed the "penultimate" rotamer library, using the quality-filtered Top240 PDBs \cite{lovell2000penultimate}. Soon afterward we updated the library using our Top500 dataset \cite{Lovell:2003uq}. This library used many file-level filters such as requiring $\le$ 1.8 resolution, a clashscore $<$ 30 \cite{Word1999}, and few backbone bond-angle outliers from Engh and Huber standards \cite{Engh1991}. \textcolor{changecolor}{If the file contained multiple identical chains, only the best-ordered one was used}. The residue-level filters required B factors $<$ 40  \textcolor{changecolor}{and occupancies of 1} for all atoms in the residue and the absence of serious steric clashes, defined as hydrogen-aware atomic overlaps $\ge$ 0.4 \AA{} \cite{Word1999}. These filters were used in order to eliminate residues with questionable justification for their given conformation, thereby increasing reliability. Specifically, the B-factor filter (the best metric available before mandatory data deposition) was meant to eliminate residues with poor electron density or other local uncertainty. However, as we realized and as Shapovalov and Dunbrack later clearly demonstrated \cite{Shapovalov:2007}, in many structures the B factors are a poor indicator of good density, primarily because they are often restrained to change only modestly between adjacent, covalently bonded atoms. Since 2008, when the wwPDB started requiring structure-factor data with all depositions, it has become feasible to do routine analysis of electron density directly.

### 2016-02-20 methods.tex
Group revisions

\section{Methods}
\subsection{\textcolor{changecolor}{Overall} Chain-level Dataset Filters}
Our previous rotamer library \cite{lovell2000penultimate} used the Top500 quality-filtered data \citep{Lovell:2003uq} as its reference dataset. Since then the number of high-resolution structures in the PDB has skyrocketed, allowing us to create a new quality-filtered, X-ray model database, the Top8000. The Top8000 was curated by assessing all PDB crystal structures as of March 29, 2011 with a protein chain of $\ge$ 38 residues at $<$ 2.0 \AA{} resolution. Hydrogens were added to each PDB file using Reduce, including Asn, Gln, and His flip corrections \citep{Word19991735}. MolProbity analysis was performed on each chain \citep{Chen:2010kx} and the results entered into a MySQL database. The chains were additionally filtered on the following criteria: chain MolProbity score $<$ 2.0, $\le$ 5\% of residues with bond length outliers ($>$ 4$\sigma$), $\le$ 5\% of residues with bond angle outliers ($>$ 4$\sigma$), and $\le$ 5\% of residues with C$\beta$ deviation outliers ($>$ 0.25\AA{}).

In order to control redundancy in the dataset, we made use of the PDB homology clusters, taken separately for 50\% sequence identity (most stringent similarity filtering), 70\%, 90\%, and 95\%. For each homology cluster, we selected the best chain based on the average of resolution and chain MolProbity score. This scoring scheme produced ties within some clusters (for $<$ 1\% of the final chain tallies); these were resolved, arbitrarily but reproducibly, by alphabetical order of PDB ID + single-character chain ID. At each homology level, a second list was chosen with the additional requirement of deposited structure-factor data. All 8 chain lists are available on GitHub \citep{Dabbish:2012} (e.g. \texttt{Top8000/Top8000-SFbest\_hom70\_pdb\_chain\_list.csv}, at \url{http://github.com/rlabduke/reference_data}). The "SF" lists are somewhat smaller, since some clusters may include no otherwise-acceptable structures with deposited data. At the 70\% similarity level optimal for most data-mining purposes, the Top8000-SFbest\_hom70 list contains 7,419 chains and the Top8000-best\_hom70 list contains 7,957 chains. \textcolor{changecolor}{We therefore call these datasets the "Top8000" as successor to the Top500.}

All rotamer statistical analysis presented here used the Top8000-SFbest\_hom70 dataset version. Deposited structure factors not only are required for electron-density based residue filtering, but they also enable examination of the model in the density map for individual examples of conformations deemed dubious or interesting. Of the 7,419 chains, 28 failed in the RSCC calculation described below, and 175 were dropped because $\le$ 20\% of their residues remained after residue-level filtering. The final rotamer dataset of 7,216 chains on GitHub \texttt{Top8000/Top8000\_rotamer\_pdb\_chain\_count.csv}.

\subsection{Residue-level Dataset Filters}
\label{sec:resfilters}
Although the chain-level filters select for structures with high overall quality, local quality varies considerably within any model. In this context, \emph{quality} refers to canonical macromolecular model validation criteria (sterics, geometry, and conformation), as well as the fit to electron density. An important aspect of this research is to include only amino-acid sidechains with sufficiently clear electron density to justify their given conformation. Our previous rotamer library was based on the Top500 dataset where only two residue-level filters were used: no clashes and no B factors $\ge$ 40 \AA\textsuperscript{2}. The B factor filter was then (in 2003) our best available proxy for electron-density quality; however, residues with dubious sidechain density still remained. To remedy this, the current dataset was also filtered by a local real-space correlation coefficient (RSCC) metric \textcolor{changecolor}{\cite{Kleywegt:ba5061}} and by local map value.

@@

To visualize the rotamer distributions, smooth contours are drawn at chosen levels using our internal programs Silk, kin2Dcont, and kin3Dcont \cite{Thesis:Word,Thesis:Davis}. For the 4-dimensional Lys and Arg cases, the 3-D plots for $\chi$1\textbf{m}, $\chi$1\textbf{t}, and $\chi$1\textbf{p} are shown separately. For validation purposes, the interpolated grid value of a given multi-$\chi$ conformation is its rotamer score. In the new system developed here, a score of $\le$0.3\% qualifies as a rotamer outlier -- its score is worse than 99.7\% of the good data. Scores $>$0.3\% and $<$ 2.0\% are considered \textit{allowed} while those $\ge$2.0\% are considered \textit{favored}, as is traditional for Ramachandran criteria \citep{Laskowski:gl0276,Lovell:2003uq}.

>>>>

\section{Methods}
\subsection{\textcolor{changecolor}{Overall} Chain-level Dataset Filters}
Our previous rotamer library distributions \cite{lovell2000penultimate} used the Top500 quality-filtered data \citep{Lovell:2003uq} as the reference dataset. Since then the number of high-resolution structures in the PDB has skyrocketed, allowing us to create a new quality-filtered, X-ray model database, the Top8000. The Top8000 was curated by assessing all PDB crystal structures as of March 29, 2011 with a protein chain of $\ge$ 38 residues at $<$ 2.0 \AA{} resolution. Hydrogens were added to each PDB file using Reduce, including Asn, Gln, and His flip corrections \citep{Word19991735}. MolProbity analysis was performed on each chain \citep{Chen:2010kx} and the results entered into a MySQL database. The chains were additionally filtered on the following criteria: chain MolProbity score $<$ 2.0, $\le$ 5\% of residues with bond length outliers ($>$ 4$\sigma$), $\le$ 5\% of residues with bond angle outliers ($>$ 4$\sigma$), and $\le$ 5\% of residues with C$\beta$ deviation outliers ($>$ 0.25\AA{}).

In order to control redundancy in the dataset, we made use of the PDB homology clusters, taken separately for 50\% sequence identity (most stringent similarity filtering), 70\%, 90\%, and 95\%. For each homology cluster, we selected the best chain based on the average of resolution and chain MolProbity score. This scoring scheme produced ties within some clusters (for $<$ 1\% of the final chain tallies); these were resolved, arbitrarily but reproducibly, by alphabetical order of PDB ID + single-character chain ID. At each homology level, a second list was chosen with the additional requirement of deposited structure-factor data. All 8 chain lists are available on GitHub \citep{Dabbish:2012} (e.g. \texttt{Top8000/Top8000-SFbest\_hom70\_pdb\_chain\_list.csv}, at \url{http://github.com/rlabduke/reference_data}). The "SF" lists are somewhat smaller, since some clusters may include no otherwise-acceptable structures with deposited data. At the 70\% similarity level optimal for most data-mining purposes, the Top8000-SFbest\_hom70 list contains 7,419 chains and the Top8000-best\_hom70 list contains 7,957 chains. \textcolor{changecolor}{We therefore call these datasets the "Top8000", as successor to the Top500.}

All rotamer statistical analysis presented here used the Top8000-SFbest\_hom70 dataset version. Deposited structure factors not only are required for electron-density based residue filtering, but they also enable examination of the model in the density map for individual examples of conformations deemed dubious or interesting. Of the 7,419 chains, 28 failed in the RSCC calculation described below, and 175 were dropped because $\le$ 20\% of their residues remained after residue-level filtering. The final rotamer dataset of 7,216 chains is listed on GitHub at \texttt{Top8000/Top8000\_rotamer\_pdb\_chain\_count.csv}.

\subsection{Residue-level Dataset Filters}
\label{sec:resfilters}
Although the chain-level filters select for structures with high overall quality, local quality varies considerably within any model. In this context, \emph{quality} refers to canonical macromolecular model validation criteria (sterics, geometry, and conformation), as well as the fit to electron density. An important aspect of this research is to include only amino-acid sidechains with sufficiently clear electron density to justify their given conformation. Our previous rotamer library was based on the Top500 dataset where only three residue-level filters were used: no clashes, no alternates, and no B factors $\ge$ 40 \AA\textsuperscript{2}. The B factor filter was then (in 2003) our best available proxy for electron-density quality; however, residues with dubious sidechain density still remained. To remedy this, the current dataset was additionally filtered by a local real-space correlation coefficient (RSCC) metric \textcolor{changecolor}{\cite{Kleywegt:ba5061}} and by local map value.

@@

To visualize the rotamer distributions, smooth contours are drawn at chosen levels using our internal programs Silk, kin2Dcont, and kin3Dcont \cite{Thesis:Word,Thesis:Davis}. For the 4-dimensional Lys and Arg cases, the 3-D plots for $\chi$1\textbf{m}, $\chi$1\textbf{t}, and $\chi$1\textbf{p} are shown separately. For validation purposes, the interpolated grid value of a given multi-$\chi$ conformation is its rotamer score. In the new system developed here, a score of $<$0.3\% qualifies as a rotamer outlier -- its score is worse than 99.7\% of the good data. Scores $\ge$0.3\% and $<$ 2.0\% are considered \textit{allowed} while those $\ge$2.0\% are considered \textit{favored}, as is traditional for Ramachandran criteria \citep{Laskowski:gl0276,Lovell:2003uq}.
